---
title: "Conditionals"
author: "A. Jesús Muñoz-Pajares"
date: "4-8 NOVEMBER 2019"
output: html_document
---


### ** Basic usage:** 
if(condition){
code
}

# Where "condition" can be everything
# that evaluates to a logical

# Conditions can be simple...
```{r eval=FALSE}
if(condition1){
code
}
```

# example: 
```{r eval=TRUE}
a<-2
if(is.numeric(a)){
	print("a is numeric.")
}
```
#  ...or more complex sentences
```{r eval=FALSE}
if(condition1 & condition2 |  condition1 & condition3){
code
}
```

```{r eval=TRUE}
# example: 
if(is.numeric(a) & a<0 | is.numeric(a) & a>0){
	print("a is numeric and it is not zero.")
}
```

# They can also be nested, that is, the "code" may contain if() functions.

```{r eval=FALSE}
  if(condition1){
  code
  }

if(condition1){
	if(condition1a){
		CODE1}
	if(condition1b){
		CODE2}
}
```
# example: 

```{r eval=TRUE}
if(is.numeric(a)){
	if(a>0){
		print("a is numeric and it is positive")
		}
	if(a<0){
		print("a is numeric and it is negative")
		}
}
```

# If condition1a and condition1b are the only two possibilities, this can be also implemented as follows:

```{r eval=FALSE}
if(condition1){
	if(condition1a){
		CODE1} else {
		CODE2}
}
```


# example: 
```{r eval=TRUE}
if(is.numeric(a)){
	if(a>0){
		print("a is numeric and it is positive")
		} else {print("a is numeric and it is negative or zero")
		}
}
```

# And also using the function ifelse, as follows:

```{r eval=FALSE}
if(condition1){
	ifelse(condition1a,CODE1,CODE2)
}
```

```{r eval=TRUE}
# example: 
if(is.numeric(a)){
	ifelse(a>0,	"a is numeric and it is positive","a is numeric and it is negative")
}
```

# Thus, in specific cases you may use the ifelse, if-else, and if-elseif-else functions

# Let's take a look on these functions with the following examples:

# *Example 1: The logarithm*
#==========================

# Imagine that you have a random number and
# you want to apply a function only
# if the number is positive

```{r eval=TRUE}
RandomNumber <- rnorm(1)
if(RandomNumber>0){
	print(log(RandomNumber))
}
```
# But maybe if it is negative you want to do
# something (e.g., print an error)
```{r eval=TRUE}
 set.seed(1)
RandomNumber <- rnorm(1)
RandomNumber
if(RandomNumber<=0){
	print("Error! Cannot compute logarithms of negative values")
}
if(RandomNumber>0){
	log(RandomNumber)
}
```
# Two if() functions can be substituted by an 
# if..else or an ifelse() function

```{r eval=TRUE}
RandomNumber <- rnorm(1)
if(RandomNumber<=0){
	"Error! Cannot compute logarithms of negative values"
} else {
	log(RandomNumber)
}
```

# Providing the result at the end:
```{r eval=TRUE}
set.seed(10)
RandomNumber <- rnorm(1)
if(RandomNumber<=0){
	Result <- "Error! Cannot compute logarithms of negative values"
} else {
	Result <- log(RandomNumber)
}
Result


RandomNumber <- rnorm(1)
ifelse(RandomNumber<=0,
	"Error! Cannot compute logarithms of negative values",
	log(RandomNumber))
```

# Example 2: Comparing with mean values
# =====================================
## Print different messages with a binary condition
## if - else

```{r eval=TRUE}
bumpus <- read.table(file="bumpus.csv",sep=",",header=T)
TL_1 <- bumpus$TL[1]
TL_avg <- mean(bumpus$TL)

if (TL_1 > TL_avg) {
    print("Bigger than the average!")
} else {
    print("Smaller or equal than the average!")
}

## when condition is not binary 
## if - else if - else
TL_sd <- sd(bumpus$TL)

if (TL_1 <= TL_avg - 1.96*TL_sd) {
    print("Below 95% of distribution")
} else { 
	if (TL_1 > TL_avg - 1.96*TL_sd & TL_1 < TL_avg + 1.96*TL_sd) {
	    print("Within the 95% distribution")
		} else {
	    print("Above the 95% distribution")}
}
```

# Example 3: Tossing a coin
# ==========================
```{r eval=TRUE}
CoinFlip  <-  sample(c("head","tail"),1)
YourChoice  <-  "head"
if(CoinFlip==YourChoice){
    print("You win!!")
}
# You will get a response only if you won

# Now let's provide a response in both cases
CoinFlip <- sample(c("head","tail"),1)
YourChoice <- "head"
if(CoinFlip==YourChoice){
"You win!!!"
}
if(CoinFlip!=YourChoice){
"Loooooseeer!! HA HA HA"
}

# But the output is lost within the code.
# We can create an object and only call at the end:
CoinFlip <- sample(c("head","tail"),1)
YourChoice <- "head"
if(CoinFlip==YourChoice){
Result <- "You win!!!"
}
if(CoinFlip!=YourChoice){
Result <- "Loooooseeer!! HA HA HA"
}
Result

# The same can be done using the if..else functions.
# In that case the second part is only evaluated if the former is FALSE
CoinFlip  <-  sample(c("head","tail"),1)
YourChoice  <-  "head"
if(CoinFlip==YourChoice) {
    Result  <-  "You win!!!"
} else {
    Result  <-  "Loooooseeer!! HA HA HA"
}
Result


# Alternatively, we can use ifelse:
CoinFlip <- sample(c("head","tail"),1)
YourChoice <- "head"
Result <- ifelse(CoinFlip==YourChoice, "You win!!","Loooooseeer!! HA HA HA")
Result
```

# Example 4: Guessing the number
# ==============================
```{r eval=TRUE}

# Randomly sample a value between 9 and 11
Random <- sample(9:11,1)

# If the remainder of dividing by 2 is 0, it is pair => 10
if(Random%%2==0){
Result <- "The number is 10"
}

# If the remainder of dividing by 3 is 0 => 9
if(Random%%3==0){
Result <- "The number is 9"
}

# If the remainder of dividing by 3 and 2 is not 0 => 11
if(Random%%2!=0 & Random%%3!=0){
Result <- "The number is 11"
}
Result

# using if-elseif-else
Random <- sample(9:11,1)
Result <- if(Random%%2==0){
"The number is 10"
} else if(Random%%3==0) {
"The number is 9"
} else {
"The number is 11"}
Result

# Using nested ifelse functions:
Random <- sample(9:11,1)
Result <- ifelse(Random%%2==0,"The number is 10",{
	ifelse(Random%%3==0,"The number is 9","The number is 11")
})
Result
```

# Example 5: rock–paper–scissors
# ==========================


# Nested if()
```{r eval=TRUE}

MyChoice <- "rock"
Rchoice <- sample(c("rock","paper","scissors"),1)
if(MyChoice=="rock"){
	if(Rchoice=="paper"){
		Result <- "Looooseeeer!!"
	}
	if(Rchoice=="scissors"){
	Result <- "You win!"
	}
	if(Rchoice=="rock"){
	Result <- "Tie!"
	}
}
if(MyChoice=="paper"){
	if(Rchoice=="scissors"){
		Result <- "Looooseeeer!!"
	}
	if(Rchoice=="rock"){
	Result <- "You win!"
	}
	if(Rchoice=="paper"){
	Result <- "Tie!"
	}
}
if(MyChoice=="scissors"){
	if(Rchoice=="rock"){
		Result <- "Looooseeeer!!"
	}
	if(Rchoice=="paper"){
	Result <- "You win!"
	}
	if(Rchoice=="scissors"){
	Result <- "Tie!"
	}
}
Result
```

# Nested ifelse() within if()

```{r eval=TRUE}

MyChoice <- "rock"
Rchoice <- sample(c("rock","paper","scissors"),1)
if(MyChoice=="rock"){
	ifelse(Rchoice=="paper",Result <- "Looooseeeer!!",ifelse(Rchoice=="scissors",Result <- "You win!",Result <- "Tie!"))
}
if(MyChoice=="paper"){
	ifelse(Rchoice=="scissors",Result <- "Looooseeeer!!",ifelse(Rchoice=="rock",Result <- "You win!",Result <- "Tie!"))
}
if(MyChoice=="scissors"){
	ifelse(Rchoice=="rock",Result <- "Looooseeeer!!",ifelse(Rchoice=="paper",Result <- "You win!",Result <- "Tie!"))
}
Result
```

# Nested ifelse() within if-elseif-else

```{r eval=TRUE}
MyChoice <- "rock"
Rchoice <- sample(c("rock","paper","scissors"),1)
if (MyChoice=="rock") {
	ifelse(Rchoice=="paper",Result <- "Looooseeeer!!",ifelse(Rchoice=="scissors",Result <- "You win!",Result <- "Tie!"))
} else if (MyChoice=="paper") {
	ifelse(Rchoice=="scissors",Result <- "Looooseeeer!!",ifelse(Rchoice=="rock",Result <- "You win!",Result <- "Tie!"))
} else {
	ifelse(Rchoice=="rock",Result <- "Looooseeeer!!",ifelse(Rchoice=="paper",Result <- "You win!",Result <- "Tie!"))
}
Result
```

# Another way
```{r eval=TRUE}
MyChoice <- "rock"
Rchoice <- sample(c("rock","paper","scissors"),1)
if(MyChoice==Rchoice)
	Result <- "Tie!"
if(MyChoice!=Rchoice)
	{
	if(MyChoice=="rock")
		ifelse(Rchoice=="paper",Result <- "Looooseeeer!!",Result <- "You win!")
	if(MyChoice=="paper")
		ifelse(Rchoice=="scissors",Result <- "Looooseeeer!!",Result <- "You win!")
	if(MyChoice=="scissors")
		ifelse(Rchoice=="rock",Result <- "Looooseeeer!!",Result <- "You win!")
	}
Result
```

rm(Result)

# Exercise:
# Find the error:

```{r eval=FALSE}
MyChoice <- "Rock"
Rchoice <- sample(c("rock","paper","scissors"),1)
if(MyChoice==Rchoice)
	Result <- "Tie!"
if(MyChoice!=Rchoice)
	{
	if(MyChoice=="rock")
		ifelse(Rchoice=="paper",Result <- "Looooseeeer!!",Result <- "You win!")
	if(MyChoice=="paper")
		ifelse(Rchoice=="scissors",Result <- "Looooseeeer!!",Result <- "You win!")
	if(MyChoice=="scissors")
		ifelse(Rchoice=="rock",Result <- "Looooseeeer!!",Result <- "You win!")
	}
Result
```

# Include a error sentence if MyChoice is not one of the expected
```{r eval=TRUE}
MyChoice <- "rock"
Rchoice <- sample(c("rock","paper","scissors"),1)
if (MyChoice=="rock") {
	ifelse(Rchoice=="paper",Result <- "Looooseeeer!!",ifelse(Rchoice=="scissors",Result <- "You win!",Result <- "Tie!"))
} else if (MyChoice=="paper") {
	ifelse(Rchoice=="scissors",Result <- "Looooseeeer!!",ifelse(Rchoice=="rock",Result <- "You win!",Result <- "Tie!"))
} else if (MyChoice=="scissors") {
	ifelse(Rchoice=="rock",Result <- "Looooseeeer!!",ifelse(Rchoice=="paper",Result <- "You win!",Result <- "Tie!"))

} else {
	print("ERROR. Available choices are paper, rock or scissors")
}
Result
```

# Include a line to convert MyChoice into lowercase
```{r eval=TRUE}
MyChoice <- "roCk"
Rchoice <- sample(c("rock","paper","scissors"),1)
if (tolower(MyChoice)=="rock") {
	Result <- ifelse(Rchoice=="paper","Looooseeeer!!",ifelse(Rchoice=="scissors","You win!","Tie!"))
} else if (tolower(MyChoice)=="paper") {
	Result <- ifelse(Rchoice=="scissors","Looooseeeer!!",ifelse(Rchoice=="rock","You win!","Tie!"))
} else if (tolower(MyChoice)=="scissors") {
	Result <- ifelse(Rchoice=="rock","Looooseeeer!!",ifelse(Rchoice=="paper","You win!","Tie!"))

} else {
	print("ERROR. Available choices are paper, rock or scissors")
}
Result
```
