---
title: "Practical 4: DEB model across elevation gradients"
author: "Urtzi Enriquez-Urzelai"
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 5)
```

# Introduction

In this practical you will use **Dynamic Energy Budget (DEB)** theory --- the
formal metabolic framework covered in today's lecture --- to model growth,
reserve dynamics, and reproduction of a large tropical lizard, the Gold Tegu
(*Tupinambis teguixin*), at three elevations in the Ecuadorian Andes.

In the previous practical (Day 2), you used a **static energy budget**:
a fixed body mass ($W_w$ = 40 g) with species-specific thermal preferences.
That model answered *how hot or active* the animal is, but could not say
*how big it grows, how much it eats, or when it reproduces*.

DEB theory adds the **ontogenetic dimension**: starting from an egg, the model
tracks three state variables

- **Structure** ($V$) --- the metabolically active tissue,
- **Reserve** ($E$) --- stored energy (yolk, fat),
- **Maturity/Reproduction buffer** ($E_H$ / $E_R$) --- the energy invested
  in development and eggs,

as functions of **temperature** and **food**. The equations are those you saw
in the lecture: the $\kappa$-rule, the von Bertalanffy growth emerging from
the surface-area--volume conflict, and the Arrhenius temperature correction.

**What you will learn:**

- Extract DEB parameters from the Add-my-Pet (AmP) database
- Run the DEB model stand-alone to predict growth and reproduction
- Couple the DEB model with the NicheMapR ectotherm model
- Compare life-history predictions across an elevation gradient
- Explore the sensitivity of growth to temperature and food

## Study sites

We use the same three sites as Practical 3:

| Site | Elev. |
|------|-------|
| Jerusalem RP | 2278 m |
| Calacalí | 2950 m |
| La Paz | 3100 m |

## The species: *Tupinambis teguixin*

The Gold Tegu is a large teiid lizard (up to ~50 cm snout--vent length,
~4 kg) from northern South America. Its DEB parameters have recently been
estimated in the AmP database. Because it is a tropical lowland species, we
expect the cooler high-elevation sites to slow its growth, delay maturity,
and reduce reproductive output.

# Setting up

```{r setup-load, message=FALSE}
library(NicheMapR)
```

Define the three sites (same coordinates as Practical 3):

```{r define-sites}
# Jerusalem RP (low)
jer_lon <- -78.261458
jer_lat <- 0.008269
jer_elev <- 2278

# Calacalí (mid)
cal_lon <- -78.517833
cal_lat <- -0.007361
cal_elev <- 2950

# La Paz (high)
paz_lon <- -79.171669
paz_lat <- -3.338333
paz_elev <- 3100
```

# Step 1: DEB parameters for *Tupinambis teguixin*

The Add-my-Pet database provides estimated DEB parameters for thousands of
species. The file `allStat.Rda` contains these estimates. Let us load it and
extract the parameters for our species.

```{r get-deb-pars, message=FALSE}
load(path.expand("~/Documents/DEB/allStat.Rda"))

species <- "Tupinambis.teguixin"

allDEB.species <- dimnames(allStat$allStat)[[1]]
allDEB.species <- allDEB.species[1:(length(allDEB.species) - 2)]

species.slot <- which(allDEB.species == species)
par.names <- unlist(labels(allStat$allStat[[species.slot]]))

for (i in seq_along(par.names)) {
  assign(par.names[i], unlist(allStat$allStat[[species.slot]][i]))
}

# The extraction above overwrites 'species' (underscore separator) ---
# restore our variable name (dot separator)
species <- "Tupinambis.teguixin"

# Some parameters may be missing --- assign sensible defaults
if (!exists("E.Hj")) E.Hj <- E.Hb
if (!exists("E.He")) E.He <- E.Hb
if (!exists("L.j"))  L.j  <- L.b

cat("Species:", species, "\n")
cat("Model type:", model, "\n")
cat("Family:", family, "| Order:", order, "\n")
cat("Max structural length L_i =", L.i, "cm\n")
cat("Max wet weight Ww_i =", round(Ww.i, 1), "g\n")
cat("Age at puberty a_p =", round(a.p), "days\n")
cat("Age at 99% of L_i a_99 =", round(a.99), "days\n")
cat("Max reproduction rate R_i =", round(R.i, 3), "eggs/day\n")
```

Here are the core DEB parameters (you learned them in the lecture):

```{r core-pars-table, echo=FALSE, results='asis'}
pars <- data.frame(
  Parameter = c("Assimilation rate {p_Am}",
                "Energy conductance v",
                "Allocation fraction k",
                "Somatic maintenance [p_M]",
                "Cost of structure [E_G]",
                "Maturity at birth E_H^b",
                "Maturity at puberty E_H^p",
                "Arrhenius temp. T_A"),
  Symbol = c("{p_Am}", "v", "k", "[p_M]", "[E_G]", "E_H^b", "E_H^p", "T_A"),
  Value = c(round(p.Am, 2), round(v, 4), round(kap, 3),
            round(p.M, 2), round(E.G, 1), round(E.Hb, 1),
            round(E.Hp, 1), T.A),
  Unit = c("J cm-2 d-1", "cm d-1", "-", "J cm-3 d-1",
           "J cm-3", "J", "J", "K")
)
knitr::kable(pars)
```

# Step 2: DEB at constant temperature

Run the DEB model at a constant 25 °C with *ad libitum* food, starting from
an egg (`start.stage = 0`). This is the classic DEB simulation --- the one that
produces a von Bertalanffy growth curve.

```{r deb-const, message=FALSE, results='hide'}
ndays <- 365 * 14         # 14 years
div <- 1
Tbs <- rep(25, ndays * div)
X <- 100                  # J/cm2, high food
Xs <- rep(X, ndays * div)
E_sm <- 350
clutchsize <- 10
mass.unit <- "g"
length.unit <- "cm"
Euler <- 0
plot <- 0
start.stage <- 0

deb <- rundeb(species = species, ndays = ndays, div = div,
              Tbs = Tbs, Xs = Xs, clutchsize = clutchsize,
              mass.unit = mass.unit, length.unit = length.unit,
              E_sm = E_sm, start.stage = start.stage,
              Euler = Euler, plot = plot)

debout <- as.data.frame(deb$debout)
debpars <- as.data.frame(deb$pars)
for (i in seq_along(deb$pars)) {
  assign(names(deb$pars[i]), deb$pars[i])
}
```

Now plot the predicted **growth in length**:

```{r plot-growth-length}
time <- seq(1, ndays) / div
plot(time, debout$length, type = "l", lwd = 2,
     xlab = "Age (days)", ylab = "Length (cm)",
     main = "Growth at 25 °C")
abline(v = which(debout$E_H > E.Hb)[1] / div, lty = 2, col = "grey")
abline(v = which(debout$E_H > E.Hp)[1] / div, lty = 2, col = "grey")
text(0, max(debout$length) * 1.02, "embryo", cex = 0.8)
text(which(debout$E_H > E.Hp)[1] / div * 0.3,
     max(debout$length), "juvenile", cex = 0.8)
text(which(debout$E_H > E.Hp)[1] / div * 1.2,
     max(debout$length), "adult", cex = 0.8)
```

And the partitioning of **mass** into structure, reserve, reproduction buffer
and gut contents:

```{r plot-growth-mass}
plot(time, debout$V, type = "l", lwd = 2, col = "darkgreen",
     ylim = c(0, max(debout$wetmass)),
     xlab = "Age (days)", ylab = "Wet mass (g)",
     main = "Mass partitioning at 25 °C")
lines(time, debout$V + debout$wetstorage + debout$wetgonad + debout$wetgut,
      lwd = 2, col = "pink")
lines(time, debout$V + debout$wetstorage + debout$wetgut,
      lwd = 2, col = "brown")
lines(time, debout$V + debout$wetstorage, lwd = 2, col = "grey")
abline(v = which(debout$E_H > E.Hb)[1] / div, lty = 2, col = "grey")
abline(v = which(debout$E_H > E.Hp)[1] / div, lty = 2, col = "grey")
legend("topleft",
       c("repro. buffer", "food in gut", "reserve", "structure"),
       lty = 1, col = c("pink", "brown", "grey", "darkgreen"), cex = 0.8)
```

> Notice how the embryo starts as mostly reserve (the yolk). After birth,
> the reserve density stabilises and growth follows von Bertalanffy dynamics.
> Once puberty is reached, the reproduction buffer begins to accumulate.

Record the key life-history predictions:

```{r lh-table, echo=FALSE, results='asis'}
birth_day <- which(debout$E_H > E.Hb)[1] / div
puberty_day <- which(debout$E_H > E.Hp)[1] / div
final_length <- tail(debout$length, 1)
final_mass <- tail(debout$wetmass, 1)

lh <- data.frame(
  Trait = c("Age at birth (days)", "Age at puberty (days)",
            "Length at 14 yr (cm)", "Mass at 14 yr (g)"),
  Value = c(round(birth_day), round(puberty_day),
            round(final_length, 1), round(final_mass, 1))
)
knitr::kable(lh)
```

# Step 3: Microclimate at the three sites

As in Practical 3, run the microclimate model for each elevation.
We obtain one year of hourly data and then repeat it 14 times for the
long-term DEB simulation.

```{r micro-runs, message=FALSE, results='hide'}
# nyears_micro <- 5

mic_jer <- micro_global(loc = c(jer_lon, jer_lat),
                        elev = jer_elev, timeinterval = 365, nyears=12)
mic_cal <- micro_global(loc = c(cal_lon, cal_lat),
                        elev = cal_elev, timeinterval = 365, nyears=12)
mic_paz <- micro_global(loc = c(paz_lon, paz_lat),
                        elev = paz_elev, timeinterval = 365, nyears=12)

# # Repeat microclimate data to cover the full simulation period
# rep_micro <- function(micro, n) {
#   for (nm in names(micro)) {
#     if (is.data.frame(micro[[nm]]) && nrow(micro[[nm]]) == 8760) {
#       micro[[nm]] <- micro[[nm]][rep(seq_len(8760), n), ]
#     } else if (is.vector(micro[[nm]]) && length(micro[[nm]]) == 8760) {
#       micro[[nm]] <- rep(micro[[nm]], n)
#     } else if (is.vector(micro[[nm]]) && length(micro[[nm]]) == 365) {
#       micro[[nm]] <- rep(micro[[nm]], n)
#     }
#   }
#   micro$ndays <- 365 * n
#   micro$nyears <- n
#   micro$timeinterval <- 365 * n
#   micro
# }
#
# mic_jer <- rep_micro(mic_jer, nyears_micro)
# mic_cal <- rep_micro(mic_cal, nyears_micro)
# mic_paz <- rep_micro(mic_paz, nyears_micro)
```

Plot the air temperature at local height for the first three days:

```{r plot-micro}
par(mfrow = c(1, 1))
plot(mic_jer$metout[1:72, "TALOC"], type = "l", col = "red", lwd = 1.5,
     ylim = c(0, 35), xlab = "Hour", ylab = "Temperature (°C)",
     main = "Air temperature at local height (first 3 days)")
lines(mic_cal$metout[1:72, "TALOC"], col = "blue", lwd = 1.5)
lines(mic_paz$metout[1:72, "TALOC"], col = "forestgreen", lwd = 1.5)
legend("topright",
       legend = c("Jerusalem (2278 m)", "Calacalí (2950 m)",
                  "La Paz (3100 m)"),
       col = c("red", "blue", "forestgreen"), lty = 1, lwd = 1.5)
```

> The temperature gradient is the same environmental driver you worked with
> in Practical 3. This time, instead of predicting body temperature alone,
> we will predict how this gradient shapes growth and reproduction.

# Step 4: Coupled DEB + ectotherm model

Now we link the DEB model to the biophysical ectotherm model. The
ectotherm model provides the **body temperature** (via heat-balance
calculations with behaviour), which feeds into the **Arrhenius temperature
correction** of the DEB rates. The DEB model, in turn, provides the
**size-dependent** heat and mass exchange --- a feedback loop.

First, define the morphological, behavioural, and thermal parameters.
These are approximate for *T. teguixin* (a large tropical heliotherm).

```{r ecto-params}
# Morphology
shape <- 3        # lizard shape
alpha_min <- 0.85
alpha_max <- 0.85
pct_wet <- 0.2

# Burrowing & behaviour
mindepth <- 2
maxdepth <- 10
shade_seek <- 1
burrow <- 1
climb <- 0
nocturn <- 0
crepus <- 0
diurn <- 1

# Thermal biology (approximate for T. teguixin)
T_pref  <- 35
CT_max  <- 42
CT_min  <- 10
T_F_min <- 28
T_F_max <- 37
T_B_min <- 20
T_RB_min <- 18
```

Prepare the DEB state to start from a hatchling (so we skip the long
embryonic development):

```{r deb-init}
V_init <- L.b^3
E_init <- E.m
E_H_init <- E.Hb
stage <- 1

# Missing temperature-response parameters
if (!exists("T.L"))  T.L  <- CT_min + 273.15
if (!exists("T.H"))  T.H  <- CT_max + 273.15
if (!exists("T.AL")) T.AL <- 5e4
if (!exists("T.AH")) T.AH <- 9e4

# Shape coefficient (not in allStat for this species)
del.M <- 0.2

# Reproduction mode --- T. teguixin is oviparous (egg-layer)
viviparous <- 0
photostart <- 3    # winter solstice
photofinish <- 2   # autumnal equinox
```

Now run the coupled model at **Jerusalem RP** (low elevation):

```{r ecto-jer, message=FALSE, results='hide', warning=FALSE}
micro <- mic_jer

ecto_jer <- ectotherm(
  DEB = 1, viviparous = viviparous,
  clutchsize = clutchsize, z.mult = 1,
  # Morphology & behaviour
  shape = shape, alpha_max = alpha_max, alpha_min = alpha_min,
  pct_wet = pct_wet, mindepth = mindepth, maxdepth = maxdepth,
  shade_seek = shade_seek, burrow = burrow, climb = climb,
  nocturn = nocturn, diurn = diurn, crepus = crepus,
  # Thermal biology
  T_F_max = T_F_max, T_F_min = T_F_min,
  T_B_min = T_B_min, T_RB_min = T_RB_min,
  CT_max = CT_max, CT_min = CT_min, T_pref = T_pref,
  # DEB parameters from allStat
  z = z, del_M = del.M, p_Xm = p.Xm, kap_X = kap.X,
  v = v / 24, kap = kap, p_M = p.M / 24, E_G = E.G,
  kap_R = kap.R, k_J = k.J / 24,
  E_Hb = E.Hb, E_Hj = E.Hj, E_Hp = E.Hp, E_He = E.He,
  h_a = h.a / (24^2), s_G = s.G,
  T_REF = T.ref, T_A = T.A,
  T_AL = T.AL, T_AH = T.AH, T_L = T.L, T_H = T.H,
  E_0 = E.0, f = f,
  d_V = d.V, d_E = d.E, d_Egg = d.E,
  mu_X = mu.X, mu_E = mu.E, mu_V = mu.V, mu_P = mu.P,
  kap_X_P = kap.P,
  n_X = c(n.CX, n.HX, n.OX, n.NX),
  n_E = c(n.CE, n.HE, n.OE, n.NE),
  n_V = c(n.CV, n.HV, n.OV, n.NV),
  n_P = c(n.CP, n.HP, n.OP, n.NP),
  n_M_nitro = c(n.CN, n.HN, n.ON, n.NN),
  L_b = L.b,
  # Initial state (hatchling)
  V_init = V_init, E_init = E_init,
  E_H_init = E_H_init, stage = stage,
  # Reproduction
  photostart = photostart, photofinish = photofinish
)

deb_jer <- as.data.frame(ecto_jer$debout)
```

**Note:** The coupled model's DEB output uses different column names than
`rundeb()`. The key columns are:

- `L_W` --- physical length (cm)
- `WETMASS` --- total wet mass (g)
- `WETGONAD` --- reproduction buffer (g)
- `WETGUT` --- gut contents (g)
- `E_R` --- reproduction buffer energy (J)
- `E_H` --- maturity energy (J)

And at **Calacalí** (mid elevation):

```{r ecto-cal, message=FALSE, results='hide', warning=FALSE}
micro <- mic_cal

ecto_cal <- ectotherm(
  DEB = 1, viviparous = viviparous,
  clutchsize = clutchsize, z.mult = 1,
  shape = shape, alpha_max = alpha_max, alpha_min = alpha_min,
  pct_wet = pct_wet, mindepth = mindepth, maxdepth = maxdepth,
  shade_seek = shade_seek, burrow = burrow, climb = climb,
  nocturn = nocturn, diurn = diurn, crepus = crepus,
  T_F_max = T_F_max, T_F_min = T_F_min,
  T_B_min = T_B_min, T_RB_min = T_RB_min,
  CT_max = CT_max, CT_min = CT_min, T_pref = T_pref,
  z = z, del_M = del.M, p_Xm = p.Xm, kap_X = kap.X,
  v = v / 24, kap = kap, p_M = p.M / 24, E_G = E.G,
  kap_R = kap.R, k_J = k.J / 24,
  E_Hb = E.Hb, E_Hj = E.Hj, E_Hp = E.Hp, E_He = E.He,
  h_a = h.a / (24^2), s_G = s.G,
  T_REF = T.ref, T_A = T.A,
  T_AL = T.AL, T_AH = T.AH, T_L = T.L, T_H = T.H,
  E_0 = E.0, f = f,
  d_V = d.V, d_E = d.E, d_Egg = d.E,
  mu_X = mu.X, mu_E = mu.E, mu_V = mu.V, mu_P = mu.P,
  kap_X_P = kap.P,
  n_X = c(n.CX, n.HX, n.OX, n.NX),
  n_E = c(n.CE, n.HE, n.OE, n.NE),
  n_V = c(n.CV, n.HV, n.OV, n.NV),
  n_P = c(n.CP, n.HP, n.OP, n.NP),
  n_M_nitro = c(n.CN, n.HN, n.ON, n.NN),
  L_b = L.b,
  V_init = V_init, E_init = E_init,
  E_H_init = E_H_init, stage = stage,
  photostart = photostart, photofinish = photofinish
)

deb_cal <- as.data.frame(ecto_cal$debout)
```

And at **La Paz** (high elevation):

```{r ecto-paz, message=FALSE, results='hide', warning=FALSE}
micro <- mic_paz

ecto_paz <- ectotherm(
  DEB = 1, viviparous = viviparous,
  clutchsize = clutchsize, z.mult = 1,
  shape = shape, alpha_max = alpha_max, alpha_min = alpha_min,
  pct_wet = pct_wet, mindepth = mindepth, maxdepth = maxdepth,
  shade_seek = shade_seek, burrow = burrow, climb = climb,
  nocturn = nocturn, diurn = diurn, crepus = crepus,
  T_F_max = T_F_max, T_F_min = T_F_min,
  T_B_min = T_B_min, T_RB_min = T_RB_min,
  CT_max = CT_max, CT_min = CT_min, T_pref = T_pref,
  z = z, del_M = del.M, p_Xm = p.Xm, kap_X = kap.X,
  v = v / 24, kap = kap, p_M = p.M / 24, E_G = E.G,
  kap_R = kap.R, k_J = k.J / 24,
  E_Hb = E.Hb, E_Hj = E.Hj, E_Hp = E.Hp, E_He = E.He,
  h_a = h.a / (24^2), s_G = s.G,
  T_REF = T.ref, T_A = T.A,
  T_AL = T.AL, T_AH = T.AH, T_L = T.L, T_H = T.H,
  E_0 = E.0, f = f,
  d_V = d.V, d_E = d.E, d_Egg = d.E,
  mu_X = mu.X, mu_E = mu.E, mu_V = mu.V, mu_P = mu.P,
  kap_X_P = kap.P,
  n_X = c(n.CX, n.HX, n.OX, n.NX),
  n_E = c(n.CE, n.HE, n.OE, n.NE),
  n_V = c(n.CV, n.HV, n.OV, n.NV),
  n_P = c(n.CP, n.HP, n.OP, n.NP),
  n_M_nitro = c(n.CN, n.HN, n.ON, n.NN),
  L_b = L.b,
  V_init = V_init, E_init = E_init,
  E_H_init = E_H_init, stage = stage,
  photostart = photostart, photofinish = photofinish
)

deb_paz <- as.data.frame(ecto_paz$debout)
```

# Step 5: Life history across elevation

Now compare the DEB outputs at the three sites. First, the body temperature
that the lizards actually experience:

```{r tc-compare}
env_jer <- as.data.frame(ecto_jer$environ)
env_cal <- as.data.frame(ecto_cal$environ)
env_paz <- as.data.frame(ecto_paz$environ)

par(mfrow = c(1, 1))
plot(env_jer$TC[1:(365 * 24)], type = "l", col = "red", lwd = 1,
     ylim = c(-5, 45), xlab = "Hour of year", ylab = "T_b (°C)",
     main = "Body temperature at three elevations")
lines(env_cal$TC[1:(365 * 24)], col = "blue", lwd = 1)
lines(env_paz$TC[1:(365 * 24)], col = "forestgreen", lwd = 1)
legend("topright",
       legend = c("Jerusalem (2278 m)", "Calacalí (2950 m)",
                  "La Paz (3100 m)"),
       col = c("red", "blue", "forestgreen"), lty = 1, lwd = 1.5)
```

Next, the **physical length** (`L_W`) through time:

```{r length-compare}
time_hr <- seq_len(nrow(deb_jer)) / 24  # hours $\rightarrow$ days

plot(time_hr, deb_jer$L_W, type = "l", col = "red", lwd = 2,
     ylim = c(0, max(deb_jer$L_W)),
     xlab = "Age (days)", ylab = "Length (cm)",
     main = "Growth across elevation")
lines(time_hr, deb_cal$L_W, col = "blue", lwd = 2)
lines(time_hr, deb_paz$L_W, col = "forestgreen", lwd = 2)
legend("topleft",
       legend = c("Jerusalem (2278 m)", "Calacalí (2950 m)",
                  "La Paz (3100 m)"),
       col = c("red", "blue", "forestgreen"), lty = 1, lwd = 1.5)
```

And the **wet mass** (`WETMASS`):

```{r mass-compare}
plot(time_hr, deb_jer$WETMASS, type = "l", col = "red", lwd = 2,
     ylim = c(0, max(deb_jer$WETMASS)),
     xlab = "Age (days)", ylab = "Wet mass (g)",
     main = "Mass growth across elevation")
lines(time_hr, deb_cal$WETMASS, col = "blue", lwd = 2)
lines(time_hr, deb_paz$WETMASS, col = "forestgreen", lwd = 2)
legend("topleft",
       legend = c("Jerusalem (2278 m)", "Calacalí (2950 m)",
                  "La Paz (3100 m)"),
       col = c("red", "blue", "forestgreen"), lty = 1, lwd = 1.5)
```

The **reproduction buffer** (`E_R`, in Joules) --- how much energy is
allocated to egg production:

```{r repro-compare}
plot(time_hr, deb_jer$E_R, type = "l", col = "red", lwd = 2,
     xlab = "Age (days)", ylab = "Reproduction buffer (J)",
     main = "Reproductive investment across elevation")
lines(time_hr, deb_cal$E_R, col = "blue", lwd = 2)
lines(time_hr, deb_paz$E_R, col = "forestgreen", lwd = 2)
legend("topleft",
       legend = c("Jerusalem (2278 m)", "Calacalí (2950 m)",
                  "La Paz (3100 m)"),
       col = c("red", "blue", "forestgreen"), lty = 1, lwd = 1.5)
```

> Drops in the reproduction buffer indicate egg-laying events. The steeper
> the rise, the faster the animal accumulates energy for reproduction.

Record the summary at the end of the simulation:

```{r summary-table, echo=FALSE, results='asis'}
final <- data.frame(
  Site = c("Jerusalem", "Calacalí", "La Paz"),
  Elevation_m = c(2278, 2950, 3100),
  Length_cm = round(c(tail(deb_jer$L_W, 1),
                      tail(deb_cal$L_W, 1),
                      tail(deb_paz$L_W, 1)), 2),
  Mass_g = round(c(tail(deb_jer$WETMASS, 1),
                   tail(deb_cal$WETMASS, 1),
                   tail(deb_paz$WETMASS, 1)), 1),
  Repro_buf_J = round(c(tail(deb_jer$E_R, 1),
                        tail(deb_cal$E_R, 1),
                        tail(deb_paz$E_R, 1)), 0)
)
knitr::kable(final)
```

> **Question:** Which elevation produces the largest tegu --- and the highest
> reproductive investment? Relate this to the temperature dependence of DEB
> rates (the Arrhenius correction you learned in the lecture). How does the
> cooler high-elevation site affect the balance between assimilation,
> maintenance, and growth?

# Challenges

Try to answer each challenge yourself first --- the code scaffolding is
provided. Solutions are given afterwards.

---

## Challenge 1: Temperature sensitivity of growth

Run the stand-alone DEB model (`rundeb()`) at four constant temperatures:
20, 25, 30 and 35 °C. Plot the growth curves (length vs age) on the same
axes. How does temperature affect the asymptotic length and the growth rate?

```{r challenge1, message=FALSE, results='hide'}
# Your code here
temps <- c(20, 25, 30, 35)
# ...
```

```{r challenge1-sol}
# --- SOLUTION Challenge 1 ---
# Run DEB at several constant temperatures
deb_list <- list()
for (i in seq_along(temps)) {
  Tbs_tmp <- rep(temps[i], ndays * div)
  deb_tmp <- rundeb(species = species, ndays = ndays, div = div,
                    Tbs = Tbs_tmp, Xs = Xs,
                    clutchsize = clutchsize,
                    mass.unit = mass.unit, length.unit = length.unit,
                    E_sm = E_sm, start.stage = start.stage,
                    Euler = Euler, plot = 0)
  deb_list[[i]] <- as.data.frame(deb_tmp$debout)
}

# Plot growth curves
plot(time, deb_list[[1]]$length, type = "l", col = "blue", lwd = 2,
     ylim = c(0, max(deb_list[[4]]$length)),
     xlab = "Age (days)", ylab = "Length (cm)",
     main = "Growth at different temperatures")
for (i in 2:length(temps)) {
  lines(time, deb_list[[i]]$length, col = i + 1, lwd = 2)
}
legend("topleft", legend = paste0(temps, " °C"),
       col = 2:5, lty = 1, lwd = 2)

# Summary
for (i in seq_along(temps)) {
  cat(temps[i], "°C: final length =",
      round(tail(deb_list[[i]]$length, 1), 2), "cm,",
      "final mass =", round(tail(deb_list[[i]]$wetmass, 1), 1), "g\n")
}
```

**Expected insight:** Higher temperature raises the assimilation and
maintenance rates via Arrhenius. Growth is faster (steeper slope), but
asymptotic length may be slightly smaller at very high temperatures because
maintenance costs increase faster than assimilation (the temperature
dependence of $[p_M]$ and $\{p_{Am}\}$ differs).

---

## Challenge 2: Food limitation

Reduce food density from $X = 100$ to $X = 10$ J/cm2. How does food
availability affect growth, reserve, and reproduction? Run `rundeb()`
at 25 °C with both food levels and compare.

```{r challenge2, message=FALSE, results='hide'}
# Your code here
```

```{r challenge2-sol}
# --- SOLUTION Challenge 2 ---
Xs_low <- rep(10, ndays * div)

deb_high <- rundeb(species = species, ndays = ndays, div = div,
                   Tbs = Tbs, Xs = Xs,
                   clutchsize = clutchsize,
                   mass.unit = mass.unit, length.unit = length.unit,
                   E_sm = E_sm, start.stage = start.stage,
                   Euler = Euler, plot = 0)
deb_low <- rundeb(species = species, ndays = ndays, div = div,
                  Tbs = Tbs, Xs = Xs_low,
                  clutchsize = clutchsize,
                  mass.unit = mass.unit, length.unit = length.unit,
                  E_sm = E_sm, start.stage = start.stage,
                  Euler = Euler, plot = 0)

dh <- as.data.frame(deb_high$debout)
dl <- as.data.frame(deb_low$debout)

par(mfrow = c(1, 2))
plot(time, dh$length, type = "l", col = "darkgreen", lwd = 2,
     ylim = c(0, max(dh$length)),
     xlab = "Age (days)", ylab = "Length (cm)",
     main = "Growth: high vs low food")
lines(time, dl$length, col = "red", lwd = 2)
legend("topleft", legend = c("X = 100", "X = 10"),
       col = c("darkgreen", "red"), lty = 1, lwd = 2)

plot(time, dh$fecundity, type = "l", col = "darkgreen", lwd = 2,
     xlab = "Age (days)", ylab = "Cumulative eggs",
     main = "Reproduction: high vs low food")
lines(time, dl$fecundity, col = "red", lwd = 2)
legend("topleft", legend = c("X = 100", "X = 10"),
       col = c("darkgreen", "red"), lty = 1, lwd = 2)
```

**Expected insight:** Lower food reduces the assimilation flux, so reserve
density drops, growth slows, asymptotic size is smaller, and reproduction
may be zero (the animal never accumulates enough energy to reproduce).

---

## Challenge 3: +3 °C warming at the low site

As in Practical 3, add 3 °C to the microclimate at Jerusalem RP and re-run
the coupled DEB+ectotherm model. Compare growth and reproduction with the
current-climate run.

```{r challenge3, message=FALSE, results='hide'}
# Your code here
```

```{r challenge3-sol}
# --- SOLUTION Challenge 3 ---
# Re-run microclimate with +3°C warming
mic_jer_warm <- micro_global(loc = c(jer_lon, jer_lat),
  elev = jer_elev, timeinterval = 365, nyears = 12, warm = 3)

micro <- mic_jer_warm
ecto_jer_warm <- ectotherm(
  DEB = 1, viviparous = viviparous,
  clutchsize = clutchsize, z.mult = 1,
  shape = shape, alpha_max = alpha_max, alpha_min = alpha_min,
  pct_wet = pct_wet, mindepth = mindepth, maxdepth = maxdepth,
  shade_seek = shade_seek, burrow = burrow, climb = climb,
  nocturn = nocturn, diurn = diurn, crepus = crepus,
  T_F_max = T_F_max, T_F_min = T_F_min,
  T_B_min = T_B_min, T_RB_min = T_RB_min,
  CT_max = CT_max, CT_min = CT_min, T_pref = T_pref,
  z = z, del_M = del.M, p_Xm = p.Xm, kap_X = kap.X,
  v = v / 24, kap = kap, p_M = p.M / 24, E_G = E.G,
  kap_R = kap.R, k_J = k.J / 24,
  E_Hb = E.Hb, E_Hj = E.Hj, E_Hp = E.Hp, E_He = E.He,
  h_a = h.a / (24^2), s_G = s.G,
  T_REF = T.ref, T_A = T.A,
  T_AL = T.AL, T_AH = T.AH, T_L = T.L, T_H = T.H,
  E_0 = E.0, f = f,
  d_V = d.V, d_E = d.E, d_Egg = d.E,
  mu_X = mu.X, mu_E = mu.E, mu_V = mu.V, mu_P = mu.P,
  kap_X_P = kap.P,
  n_X = c(n.CX, n.HX, n.OX, n.NX),
  n_E = c(n.CE, n.HE, n.OE, n.NE),
  n_V = c(n.CV, n.HV, n.OV, n.NV),
  n_P = c(n.CP, n.HP, n.OP, n.NP),
  n_M_nitro = c(n.CN, n.HN, n.ON, n.NN),
  L_b = L.b,
  V_init = V_init, E_init = E_init,
  E_H_init = E_H_init, stage = stage,
  photostart = photostart, photofinish = photofinish
)

deb_jer_warm <- as.data.frame(ecto_jer_warm$debout)

par(mfrow = c(1, 2))
plot(time_hr, deb_jer$L_W, type = "l", col = "blue", lwd = 2,
     ylim = c(0, max(deb_jer_warm$L_W)),
     xlab = "Age (days)", ylab = "Length (cm)",
     main = "Growth: current vs +3 °C")
lines(time_hr, deb_jer_warm$L_W, col = "red", lwd = 2)
legend("topleft", legend = c("Current", "+3 °C"),
       col = c("blue", "red"), lty = 1, lwd = 2)

plot(time_hr, deb_jer$E_R, type = "l", col = "blue", lwd = 2,
     xlab = "Age (days)", ylab = "Reproduction buffer (J)",
     main = "Repro. investment: current vs +3 °C")
lines(time_hr, deb_jer_warm$E_R, col = "red", lwd = 2)
legend("topleft", legend = c("Current", "+3 °C"),
       col = c("blue", "red"), lty = 1, lwd = 2)
```

**Expected insight:** Moderate warming accelerates DEB rates --- growth may
initially be faster, but higher maintenance costs could reduce ultimate size
and reproduction. At some point (if Tb exceeds the upper thermal limit),
growth would cease.

---

## Challenge 4: Compare with *Iguana iguana*

The AmP database also has the Green Iguana (*Iguana iguana*). Get its DEB
parameters, run `rundeb()` at 30 °C, and compare the growth trajectories of
these two large tropical lizards.

```{r challenge4, message=FALSE, results='hide'}
# Your code here
```

```{r challenge4-sol}
# --- SOLUTION Challenge 4 ---
iguana <- "Iguana.iguana"
iguana.slot <- which(allDEB.species == iguana)

igu_par.names <- unlist(labels(allStat$allStat[[iguana.slot]]))
for (i in seq_along(igu_par.names)) {
  assign(paste0("igu_", igu_par.names[i]),
         unlist(allStat$allStat[[iguana.slot]][i]))
}

# Set up simulation at 30 °C
Tbs_30 <- rep(30, ndays * div)
igu_Xs <- rep(100, ndays * div)

deb_igu <- rundeb(species = iguana, ndays = ndays, div = div,
                  Tbs = Tbs_30, Xs = igu_Xs,
                  clutchsize = 10,
                  mass.unit = mass.unit, length.unit = length.unit,
                  E_sm = E_sm, start.stage = 0,
                  Euler = Euler, plot = 0)

di <- as.data.frame(deb_igu$debout)

# Re-run T. teguixin at 30 °C for fair comparison
Tbs_30 <- rep(30, ndays * div)
deb_teg <- rundeb(species = species, ndays = ndays, div = div,
                  Tbs = Tbs_30, Xs = Xs,
                  clutchsize = 10,
                  mass.unit = mass.unit, length.unit = length.unit,
                  E_sm = E_sm, start.stage = 0,
                  Euler = Euler, plot = 0)
dt <- as.data.frame(deb_teg$debout)

plot(time, dt$length, type = "l", col = "goldenrod", lwd = 3,
     ylim = c(0, max(di$length)),
     xlab = "Age (days)", ylab = "Length (cm)",
     main = "T. teguixin vs I. iguana at 30 °C")
lines(time, di$length, col = "forestgreen", lwd = 3)
legend("topleft",
       legend = c("Tupinambis teguixin", "Iguana iguana"),
       col = c("goldenrod", "forestgreen"), lty = 1, lwd = 2)

cat("T. teguixin final length:", round(tail(dt$length, 1), 1), "cm\n")
cat("I. iguana  final length:", round(tail(di$length, 1), 1), "cm\n")
cat("T. teguixin final mass:", round(tail(dt$wetmass, 1), 1), "g\n")
cat("I. iguana  final mass:", round(tail(di$wetmass, 1), 1), "g\n")
```

**Expected insight:** The two species have different DEB parameters
(zoom factor, $p_{Am}$, $E_G$, etc.), reflecting their different life
histories. Which one grows faster? Which one reaches larger size? How
do the differences relate to their ecologies (arboreal vs terrestrial)?

---

# Summary

In this practical you have:

1. **Extracted DEB parameters** for *T. teguixin* from the AmP database.
2. **Run the DEB model stand-alone** to predict growth, reserve, and
   reproduction at constant temperature.
3. **Coupled DEB with the ectotherm model** to simulate life-history
   trajectories under realistic microclimates at three elevations.
4. **Compared life-history outcomes** across an elevation gradient.
5. **Explored sensitivity** to temperature, food, and warming scenarios.

DEB theory provides a **mechanistic link** between environment and
life history that static energy budgets cannot. The same core equations
($\kappa$-rule, reserve dynamics, Arrhenius correction) apply across the
animal kingdom --- from zooplankton to whales.

# References

Kooijman, S. A. L. M. (2010). *Dynamic Energy Budget Theory for Metabolic
Organisation*. Cambridge University Press.

Marques, G. M. et al. (2018). The AmP project: comparing species on the
basis of dynamic energy budget parameters. *PLOS Computational Biology*.

Kearney, M. R. & Porter, W. P. (2019). NicheMapR -- an R package for
biophysical modelling: the ectotherm and DEB models. *Ecography*.
