---
title: "Practical 4: Behavioural thermoregulation with ectoR_devel"
author: "Urtzi Enriquez-Urzelai"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 5)
```

# Introduction

The `ectoR_devel` function solves the steady-state energy balance for a single set of environmental conditions. Real animals, however, *behave* — they move between sun and shade, change posture, or retreat to water to control their body temperature.

In this practical you will **wrap `ectoR_devel` in a time loop** and add state-dependent behavioural rules. You will compare two thermoregulatory strategies:

| Strategy | Animal | Mechanism |
|----------|--------|-----------|
| **Shade-seeking** | Lizard | Move from sun to shade as body temperature rises |
| **Water immersion** | Frog | Jump into cool water when too hot, return when cooled |

# Setup

```{r setup-load, message=FALSE}
library(NicheMapR)
```

Define species parameters and thermal thresholds:

```{r species-params}
# Shared
Ww_g <- 40
alpha <- 0.85

# Lizard — dry skin, reptile geometry
liz_wet <- 5           # pct_wet (%)
liz_shape <- 3          # lizard allometric
liz_Tpref <- 34         # preferred body temperature (°C)
liz_CTmax <- 42         # critical thermal maximum (°C)

# Frog — moister skin, amphibian geometry
frog_wet <- 70          # pct_wet (%)
frog_shape <- 4         # frog allometric
frog_Tpref <- 26        # preferred body temperature (°C)
frog_CTmax <- 38        # critical thermal maximum (°C)

# Water temperature for frog refuge
T_water <- 20           # °C
```

---

# Synthetic diel microclimate

We need a full 24-hour cycle of environmental conditions. The simple models below capture the essential daily rhythm:

```{r microclimate-gen}
hour <- 0:23   # 0 = midnight

# Solar radiation (W/m²) — bell curve from 06:00 to 18:00
QSOLR <- ifelse(hour >= 6 & hour <= 18,
  1000 * sin(pi * (hour - 6) / 12), 0)

# Zenith angle (°) — 90° at horizon, ~10° at noon
Z <- ifelse(hour >= 6 & hour <= 18,
  90 - 80 * sin(pi * (hour - 6) / 12), 90)

# Air temperature (°C) — max 30 at 14:00, min 18 at 02:00
TA <- 24 + 6 * cos(2 * pi * (hour - 14) / 24)

# Ground temperature (°C)
TGRD <- TA + 2 * sin(pi * (hour - 8) / 14)
TSUBST <- TGRD

# Other variables (constant for simplicity)
TSKY <- rep(-5, 24)     # sky temperature (°C)
RH <- rep(60, 24)       # relative humidity (%)
VEL <- rep(1, 24)       # wind speed (m/s)
```

Plot the diel cycle:

```{r plot-micro, message=FALSE}
par(mfrow = c(3, 2))
plot(hour, QSOLR, type = "l", lwd = 2, col = "gold3",
  xlab = "Hour", ylab = expression("QSOLR (W/m"^2 * ")"))
plot(hour, Z, type = "l", lwd = 2, col = "skyblue",
  xlab = "Hour", ylab = "Zenith angle (°)")
plot(hour, TA, type = "l", lwd = 2, col = "firebrick",
  xlab = "Hour", ylab = "TA (°C)", ylim = range(c(TA, TGRD)))
lines(hour, TGRD, lwd = 2, col = "sienna")
legend("topright", legend = c("TA", "TGRD"), col = c("firebrick", "sienna"),
  lty = 1, lwd = 2)
plot(hour, TSKY, type = "l", lwd = 2, col = "purple",
  xlab = "Hour", ylab = "TSKY (°C)")
plot(hour, RH, type = "l", lwd = 2, col = "steelblue",
  xlab = "Hour", ylab = "RH (%)", ylim = c(0, 100))
plot(hour, VEL, type = "l", lwd = 2, col = "darkgreen",
  xlab = "Hour", ylab = "VEL (m/s)", ylim = c(0, 2))
```

# Null model -- no behaviour

What happens if the animal does nothing at all? Loop over the 24 hours and call `ectoR_devel` with fixed parameters (full sun, no shade):

```{r null-model, message=FALSE}
tc_null <- numeric(24)

for (h in 1:24) {
  tc_null[h] <- ectoR_devel(Ww_g = Ww_g, shape = liz_shape, alpha = alpha,
    pct_wet = liz_wet, TA = TA[h], TGRD = TGRD[h], TSUBST = TSUBST[h],
    TSKY = TSKY[h], VEL = VEL[h], RH = RH[h], QSOLR = QSOLR[h],
    Z = Z[h], SHADE = 0)$TC
}
```

```{r plot-null, message=FALSE}
plot(hour, tc_null, type = "o", pch = 16, lwd = 2, col = "grey40",
  xlab = "Hour", ylab = "Body temperature (°C)",
  main = "Null model -- no behaviour")
abline(h = liz_CTmax, lty = 2, col = "red")
abline(h = frog_CTmax, lty = 2, col = "blue")
text(0, liz_CTmax + 1, "Lizard CTmax", col = "red", cex = 0.7)
text(0, frog_CTmax + 1, "Frog CTmax", col = "blue", cex = 0.7)
```

Without any behaviour, both species would overheat and die. The null model reaches `r round(max(tc_null), 1)` °C — far above both CTmax values.

# Lizard — shade-seeking behaviour

The lizard thermoregulates by moving between sun and shade. At each hour, the model **tries full sun first**, and if the resulting body temperature exceeds the threshold, it seeks shade (or retreats to a burrow).

```{r lizard-model, message=FALSE}
tc_liz <- numeric(24)
shade_liz <- integer(24)

for (h in 1:24) {
  if (QSOLR[h] == 0) {
    # Night: shelter (full shade, still air)
    out <- ectoR_devel(Ww_g = Ww_g, shape = liz_shape, alpha = alpha,
      pct_wet = liz_wet, TA = TA[h], TGRD = TGRD[h], TSUBST = TSUBST[h],
      TSKY = TSKY[h], VEL = 0.3, RH = RH[h], QSOLR = 0, Z = 90,
      SHADE = 100)
    tc_liz[h] <- out$TC; shade_liz[h] <- 100
    next
  }

  # Try basking in full sun
  out <- ectoR_devel(Ww_g = Ww_g, shape = liz_shape, alpha = alpha,
    pct_wet = liz_wet, TA = TA[h], TGRD = TGRD[h], TSUBST = TSUBST[h],
    TSKY = TSKY[h], VEL = 1, RH = RH[h], QSOLR = QSOLR[h],
    Z = Z[h], SHADE = 0)

  if (out$TC > 42) {
    # Dangerously hot → retreat to burrow
    out <- ectoR_devel(Ww_g = Ww_g, shape = liz_shape, alpha = alpha,
      pct_wet = liz_wet, TA = TA[h], TGRD = TGRD[h], TSUBST = TSUBST[h],
      TSKY = TSKY[h], VEL = 0.3, RH = RH[h], QSOLR = QSOLR[h],
      Z = Z[h], SHADE = 99)
    sh <- 99
  } else if (out$TC > 38) {
    # Too warm → seek shade
    out <- ectoR_devel(Ww_g = Ww_g, shape = liz_shape, alpha = alpha,
      pct_wet = liz_wet, TA = TA[h], TGRD = TGRD[h], TSUBST = TSUBST[h],
      TSKY = TSKY[h], VEL = 1, RH = RH[h], QSOLR = QSOLR[h],
      Z = Z[h], SHADE = 85)
    sh <- 85
  } else {
    sh <- 0   # comfortable in sun
  }

  tc_liz[h] <- out$TC
  shade_liz[h] <- sh
}
```

```{r plot-lizard, message=FALSE}
par(mfrow = c(2, 1))

# TC with behaviour overlay
plot(hour, tc_null, type = "l", lwd = 1.5, col = "grey", lty = 2,
  xlab = "Hour", ylab = "Body temperature (°C)",
  main = "Lizard: shade-seeking behaviour",
  ylim = c(10, 45))
lines(hour, tc_liz, type = "o", pch = 16, lwd = 2, col = "orange")
abline(h = liz_CTmax, lty = 2, col = "red")
abline(h = liz_Tpref, lty = 3, col = "orange")
legend("topright", legend = c("Null (no behaviour)", "Lizard with behaviour",
  "Tpref", "CTmax"), col = c("grey", "orange", "orange", "red"),
  lty = c(2, 1, 3, 2), lwd = c(1.5, 2, 1, 1))

# Shade level
plot(hour, shade_liz, type = "s", lwd = 2, col = "darkgreen",
  xlab = "Hour", ylab = "SHADE (%)", main = "Shade used by lizard",
  ylim = c(0, 100))
```

The lizard basks in the morning, seeks shade during the hottest hours, and returns to sun in late afternoon. Its body temperature stays well below CTmax ($\max = `r round(max(tc_liz), 1)`$ °C).

## 5. Frog — water immersion behaviour

The frog uses a binary strategy: when it gets too hot on land, it jumps into cool water and cools instantly. Once in water, it stays until the land temperature would be comfortable again.

```{r frog-model, message=FALSE}
tc_frog <- numeric(24)
in_water <- logical(24)
in_water_prev <- FALSE    # track state across hours

for (h in 1:24) {
  if (QSOLR[h] == 0) {
    # Night on land
    out <- ectoR_devel(Ww_g = Ww_g, shape = frog_shape, alpha = alpha,
      pct_wet = frog_wet, TA = TA[h], TGRD = TGRD[h], TSUBST = TSUBST[h],
      TSKY = TSKY[h], VEL = 1, RH = RH[h], QSOLR = 0, Z = 90,
      SHADE = 100)
    tc_frog[h] <- out$TC
    in_water_prev <- FALSE
    next
  }

  # Compute what TC would be on land
  land <- ectoR_devel(Ww_g = Ww_g, shape = frog_shape, alpha = alpha,
    pct_wet = frog_wet, TA = TA[h], TGRD = TGRD[h], TSUBST = TSUBST[h],
    TSKY = TSKY[h], VEL = 1, RH = RH[h], QSOLR = QSOLR[h],
    Z = Z[h], SHADE = 0)

  if (in_water_prev && land$TC < 25) {
    # Cool enough to return to land
    tc_frog[h] <- land$TC
    in_water_prev <- FALSE
  } else if (in_water_prev || land$TC > 30) {
    # Stay in or jump to water
    tc_frog[h] <- T_water
    in_water_prev <- TRUE
  } else {
    tc_frog[h] <- land$TC
    in_water_prev <- FALSE
  }
  in_water[h] <- in_water_prev
}
```

```{r plot-frog, message=FALSE}
par(mfrow = c(2, 1))

# TC with water periods shaded
plot(hour, tc_null, type = "l", lwd = 1.5, col = "grey", lty = 2,
  xlab = "Hour", ylab = "Body temperature (°C)",
  main = "Frog: water-immersion behaviour",
  ylim = c(10, 45))
# Shade water periods
water_idx <- which(in_water)
if (length(water_idx) > 0) {
  rect(hour[water_idx] - 0.5, -5, hour[water_idx] + 0.5, 50,
    col = rgb(0.3, 0.6, 1, 0.2), border = NA)
}
lines(hour, tc_frog, type = "o", pch = 16, lwd = 2, col = "dodgerblue")
abline(h = frog_CTmax, lty = 2, col = "red")
abline(h = frog_Tpref, lty = 3, col = "dodgerblue")
abline(h = T_water, lty = 3, col = "cyan")
legend("topright", legend = c("Null", "Frog with behaviour", "Tpref",
  "CTmax", "Water"), col = c("grey", "dodgerblue", "dodgerblue", "red",
  "cyan"), lty = c(2, 1, 3, 2, 3), lwd = c(1.5, 2, 1, 1, 1))
text(0, T_water - 1, paste("T_water =", T_water), col = "cyan", cex = 0.7)

# Water/land state
plot(hour, as.integer(in_water), type = "s", lwd = 2, col = "dodgerblue",
  xlab = "Hour", ylab = "State", main = "Frog: in water?",
  yaxt = "n", ylim = c(-0.1, 1.1))
axis(2, at = c(0, 1), labels = c("Land", "Water"))
```

The frog shows a classic **sawtooth** body temperature pattern: rapid heating on land, then a sharp drop when it jumps into the water refuge. It spends `r sum(in_water)` hours in the water.

# Three-way comparison

```{r comparison, message=FALSE}
matplot(hour, cbind(tc_null, tc_liz, tc_frog), type = "o", pch = 16,
  lwd = 2, col = c("grey70", "orange", "dodgerblue"),
  xlab = "Hour", ylab = "Body temperature (°C)",
  main = "Comparison: null vs lizard vs frog")
abline(h = liz_CTmax, lty = 2, col = "red", lwd = 1.5)
abline(h = frog_CTmax, lty = 2, col = "purple", lwd = 1.5)
legend("topright", legend = c("Null", "Lizard (shade)", "Frog (water)",
  "Lizard CTmax", "Frog CTmax"),
  col = c("grey70", "orange", "dodgerblue", "red", "purple"),
  lty = c(1, 1, 1, 2, 2), lwd = 2)
```

Both behavioural strategies keep body temperature within safe limits, but they work very differently. The lizard's shade-seeking produces a smooth diel curve, while the frog's water immersion creates an abrupt sawtooth.

# Real microclimate from NicheMapR

So far we used a synthetic diel cycle. Now let us run `micro_global()` to get a realistic microclimate for Jerusalem RP (the warmest of our sites) and drive the behaviour models with real data.

## Run the microclimate model

```{r run-micro-real, message=FALSE}
jer_lon <- -78.261458
jer_lat <- 0.008269
jer_elev <- 2278

mic <- micro_global(loc = c(jer_lon, jer_lat), elev = jer_elev,
  timeinterval = 12)
```

The model extracts 12 days sampled across the year. Let us find the hottest:

```{r find-hottest}
met <- as.data.frame(mic$metout)
day_means <- tapply(met$TALOC, met$DOY, mean)
hot_doy <- as.numeric(names(which.max(day_means)))
cat("Hottest day of year:", hot_doy, "\n")
```

## Extract the day's data

```{r extract-day}
idx <- which(met$DOY == hot_doy)
hour_real <- met$TIME[idx] / 60
TA_real <- met$TALOC[idx]
QSOLR_real <- met$SOLR[idx]
Z_real <- met$ZEN[idx]
RH_real <- met$RHLOC[idx]
VEL_real <- met$VLOC[idx]
TSKY_real <- met$TSKYC[idx]

# Ground temperature from soil surface
soil_df <- as.data.frame(mic$soil)
TGRD_real <- soil_df$D0cm[idx]
```

Plot the real diel cycle:

```{r plot-real-micro, message=FALSE}
par(mfrow = c(3, 2))
plot(hour_real, TA_real, type = "l", lwd = 2, col = "firebrick",
  xlab = "Hour", ylab = "TA (°C)")
plot(hour_real, QSOLR_real, type = "l", lwd = 2, col = "gold3",
  xlab = "Hour", ylab = expression("QSOLR (W/m"^2 * ")"))
plot(hour_real, Z_real, type = "l", lwd = 2, col = "skyblue",
  xlab = "Hour", ylab = "Zenith (°)")
plot(hour_real, TGRD_real, type = "l", lwd = 2, col = "sienna",
  xlab = "Hour", ylab = "TGRD (°C)")
plot(hour_real, RH_real, type = "l", lwd = 2, col = "steelblue",
  xlab = "Hour", ylab = "RH (%)", ylim = c(0, 100))
plot(hour_real, VEL_real, type = "l", lwd = 2, col = "darkgreen",
  xlab = "Hour", ylab = "VEL (m/s)")
```

Compare with the synthetic cycle: real solar radiation peaks at ~540 W/m² (not 1000), and air temperature reaches ~31 °C. The microclimate model accounts for cloud cover, atmospheric scattering, and local topography.

## Lizard behaviour with real microclimate

Run the same shade-seeking logic but with real environmental data:

```{r lizard-real, message=FALSE}
tc_liz_real <- numeric(24)
shade_liz_real <- numeric(24)

for (h in 1:24) {
  if (QSOLR_real[h] == 0) {
    out <- ectoR_devel(Ww_g = Ww_g, shape = liz_shape, alpha = alpha,
      pct_wet = liz_wet, TA = TA_real[h], TGRD = TGRD_real[h],
      TSUBST = TGRD_real[h], TSKY = TSKY_real[h], VEL = 0.3,
      RH = RH_real[h], QSOLR = 0, Z = 90, SHADE = 100)
    tc_liz_real[h] <- out$TC; shade_liz_real[h] <- 100
    next
  }
  out <- ectoR_devel(Ww_g = Ww_g, shape = liz_shape, alpha = alpha,
    pct_wet = liz_wet, TA = TA_real[h], TGRD = TGRD_real[h],
    TSUBST = TGRD_real[h], TSKY = TSKY_real[h], VEL = 1,
    RH = RH_real[h], QSOLR = QSOLR_real[h], Z = Z_real[h], SHADE = 0)
  if (out$TC > 42) {
    out <- ectoR_devel(Ww_g = Ww_g, shape = liz_shape, alpha = alpha,
      pct_wet = liz_wet, TA = TA_real[h], TGRD = TGRD_real[h],
      TSUBST = TGRD_real[h], TSKY = TSKY_real[h], VEL = 0.3,
      RH = RH_real[h], QSOLR = QSOLR_real[h], Z = Z_real[h], SHADE = 99)
    sh <- 99
  } else if (out$TC > 38) {
    out <- ectoR_devel(Ww_g = Ww_g, shape = liz_shape, alpha = alpha,
      pct_wet = liz_wet, TA = TA_real[h], TGRD = TGRD_real[h],
      TSUBST = TGRD_real[h], TSKY = TSKY_real[h], VEL = 1,
      RH = RH_real[h], QSOLR = QSOLR_real[h], Z = Z_real[h], SHADE = 85)
    sh <- 85
  } else {
    sh <- 0
  }
  tc_liz_real[h] <- out$TC; shade_liz_real[h] <- sh
}
```

## Frog behaviour with real microclimate

```{r frog-real, message=FALSE}
tc_frog_real <- numeric(24)
in_water_real <- logical(24)
in_water_prev <- FALSE

for (h in 1:24) {
  if (QSOLR_real[h] == 0) {
    out <- ectoR_devel(Ww_g = Ww_g, shape = frog_shape, alpha = alpha,
      pct_wet = frog_wet, TA = TA_real[h], TGRD = TGRD_real[h],
      TSUBST = TGRD_real[h], TSKY = TSKY_real[h], VEL = 1,
      RH = RH_real[h], QSOLR = 0, Z = 90, SHADE = 100)
    tc_frog_real[h] <- out$TC; in_water_prev <- FALSE; next
  }
  land <- ectoR_devel(Ww_g = Ww_g, shape = frog_shape, alpha = alpha,
    pct_wet = frog_wet, TA = TA_real[h], TGRD = TGRD_real[h],
    TSUBST = TGRD_real[h], TSKY = TSKY_real[h], VEL = 1,
    RH = RH_real[h], QSOLR = QSOLR_real[h], Z = Z_real[h], SHADE = 0)
  if (in_water_prev && land$TC < 25) {
    tc_frog_real[h] <- land$TC; in_water_prev <- FALSE
  } else if (in_water_prev || land$TC > 30) {
    tc_frog_real[h] <- T_water; in_water_prev <- TRUE
  } else {
    tc_frog_real[h] <- land$TC; in_water_prev <- FALSE
  }
  in_water_real[h] <- in_water_prev
}
```

## Comparison: synthetic vs real

```{r compare-synth-real, message=FALSE}
par(mfrow = c(2, 1))

# Lizard
matplot(hour_real, cbind(tc_null, tc_liz, tc_liz_real), type = "o",
  pch = 16, lwd = 2, col = c("grey70", "orange", "darkorange"),
  xlab = "Hour", ylab = "Body temperature (°C)",
  main = "Lizard: synthetic vs real microclimate")
abline(h = liz_CTmax, lty = 2, col = "red")
legend("topright", legend = c("Null (synthetic)", "Lizard (synthetic)",
  "Lizard (real)"), col = c("grey70", "orange", "darkorange"),
  lty = 1, lwd = 2, pch = 16)

# Frog
matplot(hour_real, cbind(tc_frog, tc_frog_real), type = "o",
  pch = 16, lwd = 2, col = c("dodgerblue", "darkblue"),
  xlab = "Hour", ylab = "Body temperature (°C)",
  main = "Frog: synthetic vs real microclimate")
abline(h = frog_CTmax, lty = 2, col = "red")
abline(h = T_water, lty = 3, col = "cyan")
legend("topright", legend = c("Frog (synthetic)", "Frog (real)"),
  col = c("dodgerblue", "darkblue"), lty = 1, lwd = 2, pch = 16)
```

With real microclimate, solar radiation is lower (cloud cover, atmospheric effects), so both species stay cooler and need less behavioural intervention. The lizard rarely needs shade; the frog spends fewer hours in water.

# Challenges

## Challenge 1: Minimum shade

What is the minimum shade percentage needed at noon to keep the lizard's body temperature below 38 °C? Loop over SHADE values from 0 to 100.

```{r challenge1, message=FALSE}
# your code here
```

**Hint:** Find the index of noon (`which.max(QSOLR)`), then loop `SHADE = 0:100` using a single `ectoR_devel` call per value. Use `which.min(tc > 38) - 1` to find the answer.

## Challenge 2: Deforestation

What if the lizard has no shade at all? Set `SHADE = 0` for every daylight hour and record the TC time series. How many hours is TC above 40 °C?

```{r challenge2, message=FALSE}
# your code here
```

**Hint:** Re-run the null model but use the lizard's `pct_wet` and `shape`. Count hours with `sum(tc > 40)`.

## Challenge 3: Frog without water

Repeat the frog simulation *without* the water-immersion behaviour (just let it sit on land all day). Compare the peak TC and the number of hours above 35 °C to the frog with water access.

```{r challenge3, message=FALSE}
# your code here
```

**Hint:** Run `ectoR_devel` in a loop with frog parameters and `SHADE = 0`, same as the null model for the lizard but with frog `shape` and `pct_wet`.

## Challenge 4: Cloudy day

Set `PDIF = 0.5` (50% of solar radiation is diffuse — overcast conditions) and re-run both the lizard and frog behaviour models. How does cloud cover affect each species' thermoregulatory strategy?

```{r challenge4, message=FALSE}
# your code here
```

**Hint:** Add `PDIF = 0.5` to every `ectoR_devel` call. Does the lizard need less shade? Does the frog overheat at all?

# Bonus challenge: fixing `fluid = 1` (water mode)

The `ectoR_devel` function has a `fluid` parameter (`fluid = 1` for water) but it triggers an error because the internal root-finding function `FUN_ecto` does not define `QRESP` when `FLTYPE != 0`.

Your task: find and fix the bug, then re-run the frog simulation using `fluid = 1` instead of setting `TC = T_water`.

## Step 1: Locate the bug

```{r bonus-locate, message=FALSE}
# your code here
```

You should see that `QRESP` is only assigned inside `if (FLTYPE == 0)`. When `FLTYPE = 1`, `QRESP` is never created.

## Step 2: Patch both bugs

Two functions need fixing.

**Bug 1 — `FUN_ecto`:** `QRESP` is only assigned inside `if (FLTYPE == 0)`. Insert a default before it:

```{r bonus-patch-fun, message=FALSE}
# your code here
```

**Bug 2 — `WATERPROP`:** The function computes the thermal expansion coefficient `BETA` but does not include it in the return list. `CONV_ecto` then receives `NULL` for `BETA`, causing the Rayleigh number calculation to fail.

```{r bonus-patch-water, message=FALSE}
# your code here
```

## Step 3: Verify water mode works

```{r bonus-verify, message=FALSE}
# your code here
```

## Step 4: Re-run the frog with `fluid = 1`

In water mode, set `TA = T_water`, `QSOLR = 0`, and `VEL = 0.5`:

```{r bonus-frog-fluid, message=FALSE}
# your code here
```

```{r bonus-compare, message=FALSE}
# your code here
```

Both approaches should give nearly identical curves. The patch makes the water physics work correctly through `CONV_ecto` and `SEVAP_ecto`.

# Summary

1. Generate synthetic diel microclimate cycles
2. Wrap `ectoR_devel` in a time loop to simulate a full day
3. Implement **shade-seeking behaviour** (lizard) — continuous
   thermoregulation by adjusting SHADE
4. Implement **water-immersion behaviour** (frog) — binary
   thermoregulation by switching between land and water
5. Compare the two strategies and their effectiveness
6. *(Bonus)* Find and fix a bug in `FUN_ecto` to enable true
   water-mode physics

# Reference

Kearney, M. R. & Porter, W. P. (2017). NicheMapR — an R package for biophysical modelling: the ectotherm and dynamic energy budget models. *Ecography* 40: 1–12.
