---
title: "Data types II: dataframes, lists, and other data types"
author: "A. Kaliontzopoulou, A. Jesús Muñoz-Pajares, P. Tarroso, U. Enriquez-Urzelai"
date: "4-8 November 2019"
output: html_document
---


```{r, eval=FALSE, error=TRUE, echo=FALSE}
# Hidden code to have values to provide solutions
vec1 <- c(1, 2, 3, 4, 5)
vec3 <- rnorm(5)
vec5 <- c("male", "male", "female", "female", "female")
df1 <- data.frame(vec1, vec3, vec5)
```


## **Solutions to 2.2**

1. Row names must be unique becuase they must identify a single linw. CB112 is 
repeated.

2. 
```{r, eval=FALSE, error=TRUE, echo=TRUE}
rownames(df1) <- paste("Sample", 1:5, sep="_")
```

3. It means that the column "F1" does not exist in the dataframe.

4. The created column has lower case "f1". 

5. The vector can only have one type of data. If multiple types are found, than
the vector has the type that can hold all data. In this case, the character type
is defined to the vector because of the "male" value. When doing rbind with this
vector, we are coercing the whole table to the most simple type per column.

```{r, eval=FALSE, error=TRUE, echo=TRUE}
str(df2)
```

6. 

```{r, eval=FALSE, error=TRUE, echo=TRUE}
var1 <- rnorm(100)
var2 <- paste("Sample", 1:100) 
var3 <- as.factor(sample(c("male", "female"), 100, replace=TRUE))
my.dataframe <- data.frame(var1, var2, var3)
str(my.dataframe)

```

7. 

```{r, eval=FALSE, error=TRUE, echo=TRUE}
mean(my.dataframe[,1])
```

8. *my.list2* has named slots, while *my.list1* has only indices.

9. The difference is on the class that is returned: *my.list[[5]]* returns the 
dataframe while *my.list[5]* returns a **list** with one item that is the 
dataframe.

10. The problem here is that names are characters and we are creating a 
confusion between the first item in a list (number 1) and the second item that
is given the name "1" (character). 

```{r, eval=FALSE, error=TRUE, echo=TRUE}
my.list4[[1]]
my.list4[[2]] 
#or 
my.list4[["1"]]
```

11. The *apply* did not work with dataframe because the dataframe if firstly 
converted to matrix (*as.matrix(df1)*), and all elements will be coerced to the
most simple type that hold all data, which is *character*. So the *mean* returns
NAs. The list is not converted to a single type because each item in a list is
treated individually.

12. The dataframes are a list object with a class data.frame.

```{r, eval=FALSE, error=TRUE, echo=TRUE}
class(dt)
mode(dt)
typeof(dt)
```





