---
title: "Data types II: dataframes, lists, and other data types"
author: "A. Kaliontzopoulou, A. Jesús Muñoz-Pajares, P. Tarroso, U. Enriquez-Urzelai"
date: "4-8 November 2019"
output: html_document
---

```{r, eval=TRUE, error=TRUE, echo=FALSE}
bumpus <- read.table("bumpus.txt", sep=",", header=TRUE)
```


1. Separators are tabs
```{r, eval=TRUE, error=TRUE, echo=TRUE}
bumpus <- read.table("bumpus.txt", sep="\t", header=TRUE)
```

2. With indexing:
```{r, eval=TRUE, error=TRUE, echo=TRUE}
bumpus[1:5,]
n <- nrow(bumpus)
bumpus[(n-5):(n),]
```

3. The NAs are not correctly defined
```{r, eval=TRUE, error=TRUE, echo=TRUE}
bumpus <- read.table("bumpus.txt", sep="\t", header=TRUE, na.strings = "")
```

4. The table have counts, just divide by the number of rows
```{r, eval=TRUE, error=TRUE, echo=TRUE}
table(bumpus$survived)/nrow(bumpus)
```

```{r, eval=TRUE, error=TRUE, echo=FALSE}
bumpus.num <- bumpus[, sapply(bumpus, is.numeric)]
```



5. 
```{r, eval=TRUE, error=TRUE, echo=TRUE}
apply(bumpus.num, 2, mean)
apply(bumpus.num, 1, mean)
```

6. Use quantile with apply with prob set to 0.93
```{r, eval=TRUE, error=TRUE, echo=TRUE}
apply(X=bumpus.num, 1, quantile, probs=0.93)
```

7. Change the default argument
```{r, eval=TRUE, error=TRUE, echo=TRUE}
cor(bumpus.num, method="spearman")
```

8. The counts of young and adults. It might have also counts of an empty string
if NAs were not properly defined in the *read.table* before. 

9. The function needs and array/matrix like object and not a data.frame
```{r, eval=TRUE, error=TRUE, echo=TRUE}
prop.bumpus <- prop.table(as.matrix(bumpus.num), margin=2)
```

10. All columns must sum to 1 to be correct.

```{r, eval=TRUE, error=TRUE, echo=TRUE}
colSums(prop.bumpus)
```

```{r, eval=TRUE, error=TRUE, echo=FALSE}
sex.sur <- bumpus[,c("sex", "survived")]
```


11. Just divide by the total number of individuals. 
```{r, eval=TRUE, error=TRUE, echo=TRUE}
table(sex.sur)/nrow(sex.sur)
```

12. In this case we have to deal with missing data.
```{r, eval=TRUE, error=TRUE, echo=TRUE}
agesex <- bumpus[,c("age", "sex")]
agesex <- na.exclude(agesex)
table(agesex)/nrow(agesex)
```

13. It deals with a third dimension by treating as array
```{r, eval=TRUE, error=TRUE, echo=TRUE}
agesexsur <- bumpus[,c("age", "sex", "survived")]
agesexsur <- na.exclude(agesexsur)
table(agesexsur)

```

14. Just divide by the sum of TL
```{r, eval=TRUE, error=TRUE, echo=TRUE}
xtabs(TL ~ survived + sex, bumpus)/xtabs(~ sex + survived, bumpus)
```

15. Use *order* as indexing filter.
```{r, eval=TRUE, error=TRUE, echo=TRUE}
sorted <- bumpus[order(bumpus$TL),]
head(sorted)
```

```{r, eval=TRUE, error=TRUE, echo=FALSE}
coords <- read.table("coords.txt", sep=';', header=TRUE)
colnames(coords) <- c("sub", "lat", "lon")
```

16. check the arguments to merge
```{r, eval=TRUE, error=TRUE, echo=TRUE}
bumpus4 <- merge(bumpus, coords, by.x="subject", by.y="sub", all=TRUE)
head(bumpus4)
dim(bumpus4)
```







