---
title: "Control functions II: loops"
author: "A. Kaliontzopoulou, A. Jesús Muñoz-Pajares, P. Tarroso, U. Enriquez-Urzelai"
date: "4-8 November 2019"
output: html_document
---


## **Content Description**

In this lecture we will see how to use loops in R, including *while* and *for*
constructs.

## **Introduction**

The loops are a programming feature that is easy to grasp but may cause some
confusion at the first contact. Looping allows to automate many tasks in R and
are a basic feature of programming languages. In fact, it is so ubiquitous that
in R we have been using loops without explicitly writing the code. For instance,
every time we write an *apply*, a *rowMeans*, a *rownames* there is implicitly a
loop inside the code the iterates over data to get what we asked.

Understanding the loops constructs is a major step in the process of learning
how to program in R. We will use a simple example with the calculation of the
mean value of a group of values and adding some complexity to better understand
the concepts.

Let's define our simple vector as:

**(5, 12, 7, 13, 14, 2, 8, 13, 9, 7)**

On the first day we learned that we could use R as a simple calculating machine.
We can use that to calculate the mean:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
(5+12+7+13+14+2+8+13+9+7)/10
```

Although we get what we wanted, this is not very useful if we want to calculate 
the mean of other data, or if we have 1000 of values to type... We also learned 
that we could set up a variable and use it to hold a vector.

```{r, eval=TRUE, error=TRUE, echo=TRUE}
v <- c(5, 12, 7, 13, 14, 2, 8, 13, 9, 7)
```

We could calculate the mean by assessing elements in the vector:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
(v[1] + v[2] + v[3] + v[4] + v[5] + v[6] + v[7] + v[8] + v[9] + v[10]) / length(v)
```

This is better because we can change the original values in the vector and still
get the correct mean, but we are limited to vectors of length 10. But, if we look
to the code there is a repetition of the *v[i]*, only changing the indices. What 
if there was a way to **loop** over all indices of a vector independently of the
size?

R has two main types of loops constructs the *for* and the *while*. 


## **For loops**

The for loop has a very simple structure:


```{r, eval=FALSE, error=TRUE, echo=TRUE}
for (i in VECTOR) {
    DO_SOME_CODE
}
```

VECTOR is any vector we created that will give values to variable *i* sequentially
in each cycle. If VECTOR is (1,2,3) then *i* will assume each of those values in
three cycles:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
for (i in c(1,2,3)) {
    print(paste("iteration", i))
}
```

**Note** that we are naming the loop variable as *i*. We could give any name to
the variable. At the end of the loop, *i* will only store the last value.

```{r, eval=TRUE, error=TRUE, echo=TRUE}
i
```

but it had all values in the three cycles as shown by the *print* command. Let's
use to calculate the mean value of the vector v.

```{r, eval=TRUE, error=TRUE, echo=TRUE}
mean.val <- 0
for (i in 1:length(v)) {
    mean.val <- mean.val + v[i]
}
mean.val <- mean.val / length(v)
mean.val
```

Now we can calculate the mean of any numeric vector, independently of its length!
Of course we could have used the *mean* function built-in in R to this simple task.


---

### $\color{red}{\text{CHALLENGE}}$

1. Create a large random vector (more than 100 elements) and use a for loop to
calculate the mean value.

---


We can loop over other vectors, not only numbers of indices. Let's read again
the [bumpus](https://raw.githubusercontent.com/PhenoEvo/phenoevo.github.io/master/Intro2R/data/bumpus.txt) data:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
bumpus <- read.table("bumpus.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE)
```

We can loop over age, for instance:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
for (a in bumpus$age) {
    print(a)
}
```

The *for* is most useful when we have code that processes data instead of 
only printing. Let's calculate mean HL values for males and females. 

```{r, eval=TRUE, error=TRUE, echo=TRUE}
male <- 0
female <- 0
n.male <- 0
n.female <- 0

for (i in 1:nrow(bumpus)) {
    if (bumpus$sex[i] == 'f') {
        female <- female + bumpus$HL[i]
        n.female <- n.female + 1
    } else {
        male <- male + bumpus$HL[i]
        n.male <- n.male + 1
    }
}
male/n.male
female/n.female 
```

Instead of coding the whole mean algorithm, we could instead use the *for* loop
to create two subsets of our data depending on gender. Then we can use the *mean*
function from R to get the values.

```{r, eval=TRUE, error=TRUE, echo=TRUE}
male <- NULL
female <- NULL

for (i in 1:nrow(bumpus)) {
    if (bumpus$sex[i] == 'f') {
        female <- c(female, bumpus$HL[i])
    } else {
        male <- c(male, bumpus$HL[i])
    }
}
mean(female)
mean(male) 
```


---

### $\color{red}{\text{CHALLENGE}}$

2. Try to find a third version of the loop to get same results.

---


A less common but practical application is to perform system tasks like renaming
file, create directory structures, etc. In the next example we have 100 files 
with name as "Myfile_Today_000.txt". We want to replace the "Today" part of the
name with the current date ([example files](https://raw.githubusercontent.com/PhenoEvo/phenoevo.github.io/master/Intro2R/data/examplefiles.zip)).

```{r, eval=FALSE, error=TRUE, echo=TRUE}
folder <- "examplefiles"
files <- list.files(folder, "*", full.names=TRUE)

for (file in files) {
    newfile <- sub("Today", Sys.Date(), file)
    file.rename(file, newfile)
}    
```


---

### $\color{red}{\text{CHALLENGE}}$

3. Imagine that you need to perform 10 different runs of some analysis where, in
each run you test the influence of one parameter k from 0.0 to 0.5, every 0.05.
Create a directory structure to store your results as **Output/RUN_1/k_0.1**

---


## **While**

The difference between *for* and *while* loops is in the setup: *for* iterates
over a vector and *while* loops until a condition is *FALSE*. In practical terms,
the *while* loop is more suitable when we have to iterate until we reach some 
goal but we don't know when it will be met.

To better illustrate this let's build a small example. We will start with an 
empty vector and we will grow the vector with a random number each cycle, until
the sum of all elements in the vector is 100.

```{r, eval=TRUE, error=TRUE, echo=TRUE}
myData <- NULL
while (sum(myData) < 100) {
    item <- rnorm(1)
    myData <- c(myData, item)
}

length(myData)
```

The length of myData is huge! The reason is that *rnorm* with the default values
will return positive and negative numbers. Let's skip the negative number, only
growing the vector if the number is positive. The loops in R can be controlled
in the code with two commands: *next* that skips a cycle, and *break*, that quits
the loop. 

```{r, eval=TRUE, error=TRUE, echo=TRUE}
myData <- NULL
while (sum(myData) < 100) {
    item <- rnorm(1)
    if (item < 0) {
        next
    }
    myData <- c(myData, item)
}

```

To demonstrate the *break* command, let's create a situation where we have the 
same condition (sum < 100) but we want to stop at a maximum level of iterations,
even if we did not met the first condition.

```{r, eval=TRUE, error=TRUE, echo=TRUE}
myData <- NULL
maxIter <- 50
iter <- 0
while (sum(myData) < 100) {
    item <- rnorm(1)
    if (item < 0) {
        next
    }
    if (iter > maxIter) {
        break
    }
    myData <- c(myData, item)
    iter <- iter + 1
}

length(myData)
```


---

### $\color{red}{\text{CHALLENGE}}$

4. What is the length of myData? Why?
5. Can you recode the loop to stop when the length is 50?

---

## **Vectorization**

Knowing how to build *for* and *while* loops is a powerful skill that will help
with our analyses many times. However, we must not forget that R is a *vectorized*
language. If we can translate our logic in a loop to a vectorized version, the 
latter will, most likely, be more efficient. 

Let's see an example: How many males are in the bumpus dataset? We can code a
small for loop to count them:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
n.males <- 0
for (i in 1:nrow(bumpus)) {
    if (bumpus$sex[i] == 'm') {
        n.males <- n.males + 1
    }
}
n.males
```

However, knowing that we could operate on the vector directly with a simple 
condition would make that code a lot simpler:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
n.males <- sum(bumpus$sex == 'm')
n.males
```

We could also use *table* to get the same result as previous:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
table(bumpus$sex)
```

We can used the vectorization with subsets in R to calculate the average HL for
females:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
mean(bumpus$HL[bumpus$sex == 'f'])
```

Many times we will try to reinvent the wheel. R base has many functions 
which are extended by thousand of packages available. Many of these are extremely
well programmed and optimized for efficiency, so they are faster than to explicitly
code the loop. 

For instance instead of looping with nested if-else conditions, we can use R
built-in vectorized version, to classify samples as "big" or "small" regarding
the HL size and count them

```{r, eval=TRUE, error=TRUE, echo=TRUE}
table(ifelse(bumpus$HL < 18, "small", "big"))
```


---

### $\color{red}{\text{CHALLENGE}}$
6. Translate the average scheme in [figure](4.2_loops_average.svg) to R code (assumes a 
vector of length 5).
7. Translate the moving average scheme in [figure](4.2_loops_maverage.svg) to R code (first
create large vector). 
8. With the vector *1:100* calculate the difference between an item and the previous
for all vector. Try to find three different ways to calculate it.

---



[Solutions here.](4.2_loops_solutions.html)



