---
title: "Data types II: dataframes, lists, and other data types"
author: "A. Kaliontzopoulou, A. Jesús Muñoz-Pajares, P. Tarroso, U. Enriquez-Urzelai"
date: "4-8 November 2019"
output: html_document
---

```{r, eval=TRUE, error=TRUE, echo=FALSE}
bumpus <- read.table("bumpus.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE)
```


# **Solutions**

1. Creating vector v2 with 1000 samples

```{r, eval=TRUE, error=TRUE, echo=TRUE}
v <- runif(1000, 1, 10)
mean.val <- 0
for (i in 1:length(v)) {
    mean.val <- mean.val + v[i]
}
mean.val <- mean.val / length(v)
mean.val
```


2. Loop over bumpus$sex instead

```{r, eval=TRUE, error=TRUE, echo=TRUE}
gender <- unique(bumpus$sex)
results <- rep(NA, length(gender))

counter <- 1
for (sex in gender) {
    HL <- bumpus[bumpus$sex == sex, "HL"]
    results[counter] <- mean(HL)
    counter <- counter + 1
}
```

or even using the index instead of the counter:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
gender <- unique(bumpus$sex)
results <- rep(NA, length(gender))

for (i in 1:length(gender)) {
    HL <- bumpus[bumpus$sex == gender[i], "HL"]
    results[i] <- mean(HL)
}

names(results) <- gender

```

3. 

```{r, eval=FALSE, error=TRUE, echo=TRUE}
outdir <- "Output"
for (run in 1:10) {
    for (k in seq(0, 0.5, 0.05)) {
        newdir <- file.path(outdir, paste("Run", run, sep=""), paste("k", k, sep=""))
        dir.create(newdir, recursive=TRUE)
    }
}
```

4. The length is 51 because the iterations start at zero, so when iter > 50, we
have 51.

5. Either start at iteration 1 or use iter greater or equal to maxIter as condition.

6. 
```{r, eval=TRUE, error=TRUE, echo=TRUE}
x <- rnorm(5)

n <- length(x)
sum <- 0
i <- 1

if (n > 0) {
    while (i <= n) {
        sum = sum + x[i]
        i <- i + 1
    }
    A <- sum/n
} else {
    print("Error: vector is empty")
}
```
an equivalent but simpler version would be:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
sum = 0
if (n > 0) {
    for (i in 1:n) {
        sum = sum + x[i]
    }
    A <- sum/n
} else {
    print("Error: vector is empty")
}

```

7. 
```{r, eval=TRUE, error=TRUE, echo=TRUE}

x <- rnorm(100)

radius <- 3
n <- length(x)
ma <- rep(NA, n)
for (i in 1:n) {
    start <- i-radius
    end <- i+radius
    if (i <= radius) {
        start <- 1
    } else if (i > n - radius) {
        end <- n
    }
    ma[i] <- mean(x[start:end], na.rm=TRUE)
}
ma
```

8. 1st with explicit loop:

```{r, eval=TRUE, error=TRUE, echo=TRUE}
data.seq <- 1:100

## First - with loop
data.diff <- rep(NA, length(data.seq)-1)

for (i in 1:length(data.diff)) {
    data.diff[i] <- data.seq[i+1] - data.seq[i]
}

data.diff
```

2nd using a vectorized version:
```{r, eval=TRUE, error=TRUE, echo=TRUE}
data.diff <- data.seq[-1] - data.seq[-length(data.seq)]
data.diff
```

3rd using native R function:
```{r, eval=TRUE, error=TRUE, echo=TRUE}
data.diff <- diff(data.seq)
data.diff
```

